---
title: "Mileage of cars with Automatic or Manual Transmission"
output:
  html_document:
    df_print: paged
---
```{r load, include=FALSE}
library(dplyr, quiet=T)
library(data.table, quiet=T)
library(lubridate, quiet=T)
library('ggplot2', quiet=T)
library(plyr, quiet=T) 
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  fig.path = "figs/fig-"
)
```
# Introduction

We want to load mtcars dataset and check the data to see if it possible to answer two questions:

- "Is an automatic or manual transmission better for MPG"

- "Quantify the MPG difference between automatic and manual transmissions"

we will check the data, test the hypothesis that a transmission is better than the other and see the data on a boxplot.
Then we will do some regression analysis on the data finding the most important variables which have an influence on mpg.

# Exploratory Data Analysis

```{r}
head(mtcars)
```

The data frame is composed with 32 observations on 11 (numeric) variables.

-	mpg	Miles/(US) gallon
-	cyl	Number of cylinders
-	disp	Displacement (cu.in.)
-	hp	Gross horsepower
-	drat	Rear axle ratio
-	wt	Weight (1000 lbs)
-	qsec	1/4 mile time
-	vs	Engine (0 = V-shaped, 1 = straight)
-	am	Transmission (0 = automatic, 1 = manual)
-	gear	Number of forward gears
-	carb	Number of carburetors
First we do a t-test in order to see if alpha<0.05.
```{r}
mtcars$am = as.factor(mtcars$am)
levels(mtcars$am) = c("Automatic", "Manual")
summary(mtcars$am)

t.test(mpg~am,data=mtcars)
```

There are 19 cars with automatic transmission and 13 with manual.
Since alpha < 0.05 the null hypothesis can't be true. We search for a relation between mpg and kind of transmission. From the test above it seems Manual is better than Automatic.
We can now go further and see how the variables are correlated to mpg.

# Analysis 

We can check now the dependencies of mpg for all elements.
```{r}
fit <- lm(mpg~.-1, data=mtcars)
summary(fit)$coef
```
From the results it seems that the most important parameters are qsec, wt and am. We exclude carb, gear, vs, drat and cyl and repeat the multivariable regression.

To check if our conclusions are correct, we can see if the automatic variable selection function of R gives us the same results.
```{r}
best = step(lm(data=mtcars, mpg~.), trace=0)
summary(best)
```
We can see that both transmission and qsec gives a positive change of mpg while wt has a negative correlation.
The model explains approximatively 85% of the variance.
In the appendix the full analysis can be found.
```{r}
new2 <- mtcars[, c('mpg','wt','qsec','am')]
fit2 <- lm(mpg~.-1, data=new2)
```
```{r}
par(mfrow = c(2, 2))
plot(fit2)
```

There is not a clear trend on residuals: the model is a good approximation of the data.

# Conclusion

On average, automatic transmission is worse than manual by 2.9 mpg. However there might be situations, including weight and acceleration (1/4 mile time), in which automatic transmission perform better.

#Appendix

```{r}
boxplot(mpg~am, data=mtcars, main='Car Mileage Data', xlab='Transmission',ylab='Miles per Gallon', col=c('blue','red'), ylim=c(5,35))
```

From the above boxplot it is possible to see that Manual  transmission tend to have higher MPG than automatic as seen in the t-test data. 
We can check wether this is true for cars with different number of cylinders.
```{r}
boxplot(mpg~am*cyl, data=mtcars, col=c('gold','green'), main='Car Mileage Data', xlab='Transmission / Cylinders',ylab='Miles per Gallon', ylim=c(5,35))
```


 At 6 cilynders the difference is smaller. At 8 they are almost equivalent. however, only 2 cars have manual transmission for 8 cylinders.

```{r}
new <- mtcars[, c('mpg','wt','qsec','am','hp')]
fit <- lm(mpg~.-1, data=new)
summary(fit)$coef
new2 <- mtcars[, c('mpg','wt','qsec','am')]
fit2 <- lm(mpg~.-1, data=new2)
summary(fit2)$coef
```

```{r}
pairs(new2, panel = panel.smooth, main = "mtcars data")
```


A final graph with ggplot to see mpg, wt and am together. A similar plot of mpg, qsec and am.

```{r}
g = ggplot(new2, aes(y=mpg, x=wt, colour=factor(am)))
g = g + geom_point(size = 6, colour = "black") + geom_point(size = 4)+ geom_smooth(method = lm, se = FALSE, colour = "black")
g = g + xlab("Weight (1000 tons)") + ylab("Mileage per gallon")
g
```
```{r}
g = ggplot(new2, aes(y=mpg, x=qsec, colour=factor(am)))
g = g + geom_point(size = 6, colour = "black") + geom_point(size = 4)+ geom_smooth(method = lm, se = FALSE, colour = "black")
g = g + xlab("Weight (1000 tons)") + ylab("Mileage per gallon")
g
```
